---
title: "class11R"
author: "Katelyn Brown"
date: "2/22/2022"
output: pdf_document
---

Install BioConductor

```{r}
# install.packages("BiocManager)
# BiocManager :: install()
# BiocManager :: install("DESeq2")
library(BiocManager)
library(DESeq2)
```

Upload the count data and meta files.

```{r}
counts <- read.csv("https://bioboot.github.io/bimm143_W18/class-material/airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("https://bioboot.github.io/bimm143_W18/class-material/airway_metadata.csv")
head(counts)
head(metadata)
```

```{r}
control <- metadata[metadata[,"dex"]=="control",]
control.counts <- counts[ ,control$id]
control.mean <- rowSums( control.counts )/4 
head(control.mean)
```

> Q3. How would you make the above code in either approach more robust?

You can make the above code more robust by using summary() instead of just finding the mean.

> Q4. Follow the same procedure for the treated samples (i.e. calculate the mean per gene across drug treated samples and assign to a labeled vector called treated.mean)

```{r}
treated <- metadata[metadata[,"dex"]=="treated",]
treated.counts <- counts[ ,treated$id]
treated.mean <- rowSums( treated.counts )/4 
head(treated.mean)
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

