---
title: "class12R"
author: "Katelyn Brown PID 15891811"
date: "2/24/2022"
output: pdf_document
---

Upload data.

```{r setup, message = FALSE}
library(DESeq2)
library(ggplot2)
library(AnnotationDbi)
library(org.Hs.eg.db)
```

## 1. Input the counts and metadata files.

```{r input}
metaFile <- "GSE37704_metadata.csv"
countFile <- "GSE37704_featurecounts.csv"
colData <- read.csv(metaFile, row.names = 1)
head(colData)
countData <- read.csv(countFile, row.names = 1)
head(countData)
```

> Q. Remove first column of countData.

```{r}
countData <- as.matrix(countData[,-1])
head(countData)
```

> Q. Check if colData matches the row names.

```{r}
all(colData$id == colnames(countData))
```

> Q. Remove genes that have 0 read count across all samples.

```{r filter}
counts <- countData[rowSums(countData) != 0,]
head(counts)
```

## 2. Run DESeq.

```{r deseq}
dds <- DESeqDataSetFromMatrix(countData = counts, colData = colData, design = ~condition)
dds <- DESeq(dds)
dds
```

> Q. Call the summary() function on your results to get a sense of how many genes are up or down-regulated at the default 0.1 p-value cutoff.

```{r}
res <- results(dds, contrast = c("condition", "hoxa1_kd", "control_sirna"))
summary(res)
```

## 3. Add annotation.

> Q. Use the mapIDs() function multiple times to add SYMBOL, ENTREZID and GENENAME annotation to our results by completing the code below.

```{r annotation}
columns(org.Hs.eg.db)
res$symbol <- mapIds(org.Hs.eg.db, keys = row.names(res), keytype = "ENSEMBL", column = "SYMBOL", multiVals = "first")
head(res$symbol)
res$entrez <- mapIds(org.Hs.eg.db, keys = row.names(res), keytype = "ENSEMBL", column = "ENTREZID", multiVals = "first")
head(res$entrez)
res$genename <- mapIds(org.Hs.eg.db, keys = row.names(res), keytype = "ENSEMBL", column = "GENENAME", multiVals = "first")
head(res$genenames)
```

## 4. Make volcano plot.

Common summary figure that gives an overview of the results.

```{r}
plot(res$log2FoldChange, -log(res$padj))
```

Try ggplot.

```{r}
tmp <- as.data.frame(res)
tmp$fc <- abs(res$log2FoldChange) > 2

ggplot(tmp) + aes(log2FoldChange, -log(padj), col = tmp$fc) + geom_point()
```

Use Enhanced Volcano.

```{r}
library(EnhancedVolcano)
EnhancedVolcano(tmp, lab = tmp$symbol, x = 'log2FoldChange', y = 'pvalue')
```

Save results.

```{r}
write.csv(res, file = "deseq_results.csv")
```

## 5. Pathway analysis.

Try to bring back the biology and help with the interpretation of the results. Try to answer the question: which pathways and functions feature heavily in our differentially expressed genes?

```{r}
library(pathview)
library(gage)
library(gageData)
```

```{r}
data(kegg.sets.hs)
data(sigmet.idx.hs)
```

```{r}
kegg.sets.hs <- kegg.sets.hs[sigmet.idx.hs]
head(kegg.sets.hs, 3)
```

Make a vector and use gage.

```{r}
foldchanges <- res$log2FoldChange
names(foldchanges) <- res$entrez
head(foldchanges)
keggres <- gage(foldchanges, gsets = kegg.sets.hs)
attributes(keggres)
```

Look at down and up pathways. 

```{r}
head(keggres$less)
head(keggres$greater)
```

Use pathview to visualize results from cell cycle.

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa04110")
```

Look at top 5 pathways.

```{r}
keggrespathways <- rownames(keggres$greater)[1:5]

keggresids <- substr(keggrespathways, start=1, stop=8)
keggresids
```

Use pathview to visualize top 5 pathways.

```{r}
pathview(gene.data=foldchanges, pathway.id=keggresids, species="hsa")
```

> Q. Can you do the same procedure as above to plot the pathview figures for the top 5 down-regulated pathways?

```{r}
keggresdown <- rownames(keggres$less)[1:5]
keggresdownids <- substr(keggresdown, start = 1, stop = 8)
keggresdownids

pathview(gene.data = foldchanges, pathway.id = keggresdownids, species = "hsa")
```


### Gene Ontology

```{r}
data(go.sets.hs)
data(go.subs.hs)
gobpsets <- go.sets.hs[go.subs.hs$BP]
gobpres <- gage(foldchanges, gsets=gobpsets, same.dir=TRUE)
lapply(gobpres, head)
```

### Reactome

```{r}
sig_genes <- res[res$padj <= 0.05 & !is.na(res$padj), "symbol"]
print(paste("Total number of significant genes:", length(sig_genes)))
```

```{r}
write.table(sig_genes, file="significant_genes.txt", row.names=FALSE, col.names=FALSE, quote=FALSE)
```

> Q. What pathway has the most significant “Entities p-value”? Do the most significant pathways listed match your previous KEGG results? What factors could cause differences between the two methods?

The endosomal/vacuolar pathway has the most significant entities p-value, and the cell cycle/mitotic pathway has a significant entities p-value. The cell cycle pathway matches the KEGG results. Differing databases can cause the differences between the methods.