---
title: "class14R"
author: "Katelyn Brown A15891811"
date: "3/3/2022"
output:
  pdf_document:
    latex_engine: xelatex
---

# Read our vaccination data.

Downloaded the most recently dated "Statewide COVID-19 Vaccines Administered by ZIP Code" CSV. 

```{r}
vax <- read.csv("covid19vaccinesbyzipcode_test.csv")
head(vax)
```

> Q1. What column details the total number of people fully vaccinated?

The persons_fully_vaccinated column details the total number of people fully vaccinated.

> Q2. What column details the Zip code tabulation area?

The zip_code_tabulation_area column details the Zip code tabulation area. 

> Q3. What is the earliest date in this dataset?

The earliest date in the dataset is 2021-01-05.

> Q4. What is the latest date in the dataset?

```{r}
vax$as_of_date[nrow(vax)]
```

The latest date in the dataset is 2022-03-01. 

```{r}
# install.packages(skimr)
library(skimr)
skimr::skim(vax)
```

> Q5. How many numeric columns are in this dataset?

There are 9 numeric columns in the dataset.

> Q6. Note that there are “missing values” in the dataset. How many NA values there in the persons_fully_vaccinated column?

There are 18338 NA values in the persons_fully_vaccinated column.

> Q7. What percent of persons_fully_vaccinated values are missing (to 2 significant figures)?

17% of persons_fully_vaccinated is missing.

> Q8. Why might this data be missing?

This data could be missing due to lack of reporting, or possibly lack of access. Some counties may also not be reporting vaccination rates.

```{r}
# install.packages("lubridate")
library(lubridate)
```

```{r}
today()
age <- today() - ymd("2001-11-16")
age
time_length(age, "year")
```

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
today() - vax$as_of_date[1]
```

> Q9. How many days have passed since the last update of the dataset?

```{r}
today() - vax$as_of_date[nrow(vax)]
```

```{r}
# Determine how many days the dataset spans
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

2 days have passed since the last update of the dataset.

> Q10.  How many unique dates are in the dataset (i.e. how many different dates are detailed)?

```{r}
length(unique(vax$as_of_date))
```

There are 61 unique dates in the dataset.

# Working with Zip Codes

First, download the zipcodeR package and load it in the library.

```{r}
# install.packages("zipcodeR")
library(zipcodeR)
```

```{r}
geocode_zip('92037')
zip_distance('92037', '92109')
reverse_zipcode(c('92037', '92109'))
```

```{r}
zipdata <- reverse_zipcode(vax$zip_code_tabulation_area)
zipdata
```

Filter to only work with San Diego county. 

```{r}
# Base R
dim(vax[vax$county == "San Diego", ])
# Dplyr
library(dplyr)
sd <- filter(vax, county == "San Diego")
dim(sd)
```

> Q11. How many distinct zip codes are listed for San Diego County?

```{r}
length(unique(sd$zip_code_tabulation_area))
```

There are 107 unique zip codes in San Diego County.

> Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset?

```{r}
sd[which.max(sd$age12_plus_population), "zip_code_tabulation_area"]
```

92154 has the largest 12+ population in the dataset.

> Q13. What is the overall average “Percent of Population Fully Vaccinated” value for all San Diego “County” as of “2022-03-01”?

```{r}
sd$as_of_date[nrow(sd)]
```

```{r}
# Filter to day
sd.latest <- filter(sd, as_of_date == "2022-03-01")
mean(sd.latest$percent_of_population_fully_vaccinated, na.rm = TRUE)*100
summary(sd.latest$percent_of_population_fully_vaccinated, na.rm = T)
```

70.53% of San Diego county is fully vaccinated as of 2022-03-01.

> Q14. Using either ggplot or base R graphics make a summary figure that shows the distribution of Percent of Population Fully Vaccinated values as of “2022-03-01”?

```{r}
library(ggplot2)
ggplot(sd.latest) + aes(x = sd.latest$percent_of_population_fully_vaccinated) + geom_histogram() + labs(x = "Percent of Population Fully Vaccinated in Zip Code Area", y = "Count (Zip Code Area)", title = "Histogram of Vaccination Rates Across San Diego County")
```

Filter to focus on UCSD.

```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```

> Q15. Using ggplot make a graph of the vaccination rate time course for the 92037 ZIP code area.

```{r}
baseplot <- ggplot(ucsd) +
  aes(x = as_of_date, y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) + theme_linedraw() +
  ylim(c(0,1)) +
  labs(x = "Date", y="Percent Vaccinated", title = "Vaccination Rate in La Jolla, CA 92037")
baseplot
```

> Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-03-01”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
# Subset other CA zip codes with populations as big as 92037
vax.36 <- filter(vax, age5_plus_population > 36144 & as_of_date == "2022-03-01")
head(vax.36)
# Find mean of this data
vax.36mean <- mean(vax.36$percent_of_population_fully_vaccinated, na.rm = TRUE)
vax.36mean
# Add mean to baseplot
baseplot + geom_hline(yintercept = vax.36mean, linetype = 2, alpha = 0.5, color = "red")
```

> Q17. What is the 6 number summary (Min, 1st Qu., Median, Mean, 3rd Qu., and Max) of the “Percent of Population Fully Vaccinated” values for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2022-03-01”?

```{r}
summary(vax.36$percent_of_population_fully_vaccinated)
```

> Q18. Using ggplot generate a histogram of this data.

```{r}
ggplot(vax.36) + aes(percent_of_population_fully_vaccinated) + geom_histogram() + xlim(c(0,1))
```

> Q19. Is the 92109 and 92040 ZIP code areas above or below the average value you calculated for all these above?

```{r}
vax %>% filter(as_of_date == "2022-03-01") %>%  
  filter(zip_code_tabulation_area=="92109") %>%
  select(percent_of_population_fully_vaccinated)
vax %>% filter(as_of_date == "2022-03-01") %>%  
  filter(zip_code_tabulation_area=="92040") %>%
  select(percent_of_population_fully_vaccinated)
```

92109 and 92040 zip codes are below the average of 0.735 that was calculated previously.

> Q20. Finally make a time course plot of vaccination progress for all areas in the full dataset with a age5_plus_population > 36144.

```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)
ggplot(vax.36.all) +
  aes(as_of_date,
      percent_of_population_fully_vaccinated, 
      group = zip_code_tabulation_area) +
  geom_line(alpha = 0.2, color = "blue") +
  ylim(0, 1) +
  labs(x = "Date", y = "Percent Vaccinated",
       title = "Vaccination Rate Across California",
       subtitle = "Only areas with populations above 36k are shown.") +
  geom_hline(yintercept = vax.36mean, linetype = 2)
```

