---
title: "Class10R"
author: "Katelyn Brown"
date: "2/17/2022"
output: pdf_document
---

## Examine 1000 Genome Data

> Q5. What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are
homozygous for the asthma associated SNP (G|G)?

```{r}
# Read genotype file ofrom Ensembl
mxl <- read.csv("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
mxl
```

```{r}
# Use table() to answer question
mxl$Genotype..forward.strand.
table(mxl$Genotype..forward.strand.)/nrow(mxl)
```

14% of mxl are homozygous for the asthma associated SNP (G|G).

Examine GBR data set for the same SNP to compare populations.

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
gbr
```

```{r}
table(gbr$Genotype..forward.strand.)/nrow(gbr)
```

About 29% of gbr are homozygous for the same SNP.

## Expression by Genotype Analysis

I want to read my RNA-Seq expression results into R. This file is not a CSV but rather has fields separated by space. Because of this, use `read.table()`

```{r}
x <- read.table("rs8067378_ENSG00000172057.6.txt")
head(x)
```

> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

> Is the mean expression different based on genotype?

```{r}
table(x$geno)
summary(x[x$geno == "A/A",]$exp)
summary(x[x$geno == "A/G",]$exp)
summary(x[x$geno == "G/G",]$exp)
```

Based on the summary statistics of each genotype, the median expression levels are 31.25 for A/A, 25.065 for A/G, and 20.074 for G/G.

## Make a summary overview figure.

Use ggplot to make a boxplot with data.

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)
```

```{r}
ggplot(x) + aes(x = geno, y = exp, fill = geno) + geom_boxplot(notch = TRUE) + labs(title = "Genotype Versus Gene Expression", x = "Genotype", y = "Gene Expression")
```

Based on the boxplot, the A/A genotype is correlated with a higher gene expression level compared to both A/G and G/G.